Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2019-12-10 man/ecosystems/ms_ecosystems.Rmd
update ms

---
title: "Oak resilience to drought and land-use show how the ecological and geographical rear-edges do not necessarily meet today"
output:
  word_document: 
    fig_caption: yes
    fig_height: 6
    highlight: null
    reference_docx: ../../templates/template_ms.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecosystems.csl
---

@@

In the present study, we used remote-sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and radial tree growth (as a proxy for secondary growth) of *Quercus pyrenaica* Willd. (*melojo oak*) in southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change. We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) over the long term (in the last few decades). Our main hypothesis is that these relict populations driven by historical land use are vulnerable to climate change at their present climatic (either elevational or latitudinal) rear edge, and hence will show low resilience to extreme drought along climatic gradients at a small scale within the rear edge. To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests at their  present geographical rear edge; (*ii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iii*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale. For the latter task, we characterized the variability in the forest response to drought within the current geographical rear edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of *melojo* oak forests differ in their resistance, resilience and recovery to extreme drought events.

@@

The Iberian Peninsula underwent several extreme drought episodes in the last three decades (*e.g*. 1995, 1999, 2005, 2012) [@VicenteSerrano2014]. The 2005 and 2012 drought events have been documented as being among the worst in recent decades for the southern Iberian Peninsula [@Pascoa2017], and they were characterized as extreme drought in our climatic data (Figure S1; Table S3). We focused on these two drought events because they were included in the period having remote-sensing information of high spatial resolution (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize the results for 2005 and 2012 and to evaluate forest resilience to drought over a longer term (see Table S3). A drought event was identified using the SPEI 12-months scale, following a procedure similar to the one proposed by @Spinoni2015. We used 0.5º grid cells covering Sierra Nevada taken from the Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values fall below that threshold for at least two consecutive months. For each drought event, we computed: the *duration*, *i.e.* number of consecutive months with the SPEI lower than a certain threshold; the *severity*, *i.e.* the sum of the SPEI values (absolute values) during the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.

@@

Greenness and tree growth were more affected by drought events in drier northern populations than in wetter southern oak populations. For example, the northern site showed higher browning intensity than did the southern sites during the 2005 drought event, and the stronger correlations of tree-growth with SPEI (hydrological and summer) in the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdo2014]. It is well known that tree growth and tree responses to drought are site-dependent [*e.g.* soil features, tree competition, etc.; @Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM]. Trees at CA-High registered higher BAI than those located at lower elevations (CA-Low and SJ; Figure 4). This shows the high variability in the response to climate exhibited along a narrow gradient, which is especially noteworthy for southern sites, as these lie very close to each other and overall both are considered to constitute the rear edge for the species.

@@

### Relict oaks show high resilience to recent drought events and long-term climatic variability 
Despite the severe drought events in recent decades (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* for the last 16 years, which is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather wet decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto]. For tree growth, positive trends also appeared in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends have been described for this species along its distribution range only at high-elevation wet and cold sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold-limited high-elevation sites [@Salzer2009; @GeaIzquierdo2014]. Importantly, for rear edges threatened by climate change, negative growth trends would have been expected, as shown for some temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017].

@@

Second, these resilience responses of oak forest to drought events are not spatially homogeneous throughout the mountain range, due to differences in ecological conditions and/or past-management legacies. In fact, there was much small-scale variability in the response to climate within the rear edge that we had *a priori* considered in our study. Furthermore, we even found positive effects of climate change in certain stands, as discussed, in disagreement with our hypothesis of expecting oak vulnerability in the geographical rear edge studied. This suggests that the rear edge therefore needs to be redefined in [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear edge) and the ecological (limiting) rear edge of species.

>>>>

---
title: "Oak resilience to drought and land use show how the ecological and geographical rear edges do not necessarily meet today"
output:
  word_document: 
    fig_caption: yes
    fig_height: 6
    highlight: null
    reference_docx: ../../templates/template_ms.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecosystems.csl
---

@@

In the present study, we used remote-sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and radial tree growth (as a proxy for secondary growth) of *Quercus pyrenaica* Willd. (Pyrenean oak, *melojo*) in southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change. We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) over the long term (in the last few decades). Our main hypothesis is that these relict populations driven by historical land use are vulnerable to climate change at their present climatic (either elevational or latitudinal) rear edge, and hence will show low resilience to extreme drought along climatic gradients at a small scale within the rear edge. To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests at their  present geographical rear edge; (*ii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iii*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale. For the latter task, we characterized the variability in the forest response to drought within the current geographical rear edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of *melojo* oak forests differ in their resistance, resilience and recovery to extreme drought events.

@@

The Iberian Peninsula underwent several extreme drought episodes in the last three decades (*e.g*. 1995, 1999, 2005, 2012) [@VicenteSerrano2014]. The 2005 and 2012 drought events have been documented as being among the worst in recent decades for the southern Iberian Peninsula [@Pascoa2017], and they were characterized as extreme drought in our climatic data (Figure S1; Table S3). We focused on these two drought events because they were included in the period having remote-sensing information of high spatial resolution (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize the results for 2005 and 2012 and to evaluate forest resilience to drought over a longer term (see Table S3). A drought event was identified using the SPEI 12-months scale, following a procedure similar to the one proposed by @Spinoni2015. We used 0.5º grid cells covering Sierra Nevada taken from the Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values fall below that threshold for at least two consecutive months. For each drought event, we computed: the *duration* as the number of consecutive months with the SPEI lower than a certain threshold; the *severity* as the sum of the absolute SPEI values during the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.

@@

Greenness and tree growth were more affected by drought events in drier northern populations than in wetter southern oak populations. For example, the northern site showed higher browning intensity than did the southern sites during the 2005 drought event, and the stronger correlations of tree-growth with SPEI (hydrological year and summer) in the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdo2014]. It is well known that tree growth and tree responses to drought are site-dependent [*e.g.* soil features, tree competition, etc.; @Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM]. Trees at CA-High registered higher BAI than those located at lower elevations (CA-Low and SJ; Figure 4). This shows the high variability in the response to climate exhibited along a narrow gradient, which is especially noteworthy for southern sites, as these lie very close to each other and overall both are considered to constitute the rear edge for the species.

@@

### Relict oaks show high resilience to recent drought events and long-term climatic variability 
Despite the severe drought events in recent decades (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* for the last 16 years, which is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather wet decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto]. For tree growth, positive trends also appeared in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends have been described for this species along its distribution range only at high-elevation wet and cold sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming for the species at cold-limited high-elevation sites [@Salzer2009; @GeaIzquierdo2014]. Importantly, for rear edges threatened by climate change, negative growth trends would have been expected, as shown for some temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017].

@@

Second, these resilience responses of oak forest to drought events are not spatially homogeneous throughout the mountain range, due to differences in ecological conditions and/or past-management legacies. In fact, there was much small-scale variability in the response to climate within the rear edge that we had *a priori* considered in our study. Furthermore, we even found positive effects of climate change in certain stands, as discussed, in disagreement with our hypothesis of expecting oak vulnerability in the geographical rear edge studied. This suggests that the rear edge therefore needs to be redefined [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear edge) and the ecological (limiting) rear edge of species.

### 2019-12-10 man/figures/0_figures_ecosystems.Rmd
update ms

**Figure 1**. Distribution of *Quercus pyrenaica* forests in the Iberian Peninsula (a) and in Sierra Nevada mountain range (b). Different colours indicate oak population cluster's identified in Sierra Nevada [@PerezLuque2015]. For each population, a grid with the MODIS pixels is shown (see material and methods). Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d). Colour orthophotography of 2009 from Regional Ministry of the Environment.

@@

**Figure 2.** ***a)*** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error. See main text for details on EVI calculation. ***b)*** Percentage of pixels showing browning, greening or no-changes during the 2005 and 2012 drought events according to EVI standardized anomalies. See main text for an explanation of greening and browning.

@@

**Figure 3.** Basal Area Increment (BAI) chronologies of *Q. pyrenaica* for northern population (SJ; *red*) and southern ones: low-elevation (CA-Low; *green*) and high-elevation (CA-High, *blue*) sites. Shading areas correspond to standard error of the mean. Number of series is displayed in the upper plot. We only show years replicated with # series > 5. Linear trends since 1975 are shown all sites (numbers indicate $r^2$ values; asterisks indicate significant linear trend, *p < 0.001*).

@@

**Figure 4.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods). Note that y-axes do not correspond in all of the three panels for the sake of clarity. Error bars indicate standard error.

@@

**Figure 5.** Resilience metrics of tree-growth for eigth severe drought events since 1950 (see main text for details) as a function of drought severity. *Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*). Points indicate resilience metrics for oak populations: SJ (*red*), CA-High (*blue*) and CA-Low (*green*). Resilience metrics were computed for each population (sample depth > 10) and drought event. Gray lines represent overall relationships for each Resilience metrics.

@@

**Figure 6.** Comparison of the response of  *Q. pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (a) and tree-growth (b). For EVI we compared northern populations (*red circle*) with southern ones (*green circle*). For BAI we compared the more xeric northern population (San Juan, SJ; *red circle*) with the two southerns populations: Cáñar-High (CA-High; *blue circle*) and Cáñar-Low (CA-Low; *green circle*). Different letters indicate significant *post hoc* differences between groups (see material and methods for details).

@@

**Figure 7.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q. pyrenaica* and monthly climatic data: precipitation and 6-month SPEI (a), minimun (b) and maximun (c) temperatures. *green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High). Asteriks indicate significant ($P < 0.05$) correlation coefficients.

@@

**Figure S1.** ***a)*** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017. Points represent mean and errorbars standard error. *Black* line indicates mean for the whole period (585 mm). *Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively). *Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively). Years with average values below -1SD are labelled. Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET). ***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*). The boxplot representing the average from 1950-2015 period. Data from meteorological station Granada, Base Aérea. ***b)*** Drought severity in Sierra Nevada for the 1901-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI). Data from Global SPEI database (http://spei.csic.es/database.html). We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada. Horizontal gray bars indicate 2005 and 2012 year.

@@

**Figure S2.** ***a)*** Residual tree-ring chronologies obtained for the *Q. pyrenaica* sites. Dashed red lines indicate the start of the reliable period (EPS > 0.85). Dotted black lines showing the severe drought years identified in our climatic data (Table S3 and Figure S1). ***b)*** Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis). Data for number of trees > 2 is shown.

@@

**Figure S3.** ***a)*** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years). Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated. Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot`. ***b)*** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).

@@

##### 
**Figure S4.** EVI annual profile (average of the period 2000-2016) for *Q. pyrenaica* forests in Sierra Nevada and drought events. Horizontal bars correspond to the most severe droughts for Sierra Nevada since 1900 (computed as in Table S3). Their position indicate the start and end months of each drought event. Bars length show the duration of the drought event (number of consecutive months with SPEI lower than -1.28, see [@Pascoa2017]).

@@

**Figure S5.** Resilience metrics of the tree-growth for severe drought events since 1950 (excluding 1995 drought event). *Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*). Points indicate resilience metrics for oak populations: SJ (*blue*), CA-High (*red*) and CA-Low (*green*). Resilience metrics were computed for each population (sample depth > 10) and drought event. Gray line represents overall relationship for each Resilience metrics.

>>>>

**Figure 1**. Distribution of *Quercus pyrenaica* forests in the Iberian Peninsula (a) and in Sierra Nevada mountain range (b). Different colors indicate oak-population clusters identified in Sierra Nevada [@PerezLuque2015]. For each population, a grid with the MODIS pixels is shown (see Material and methods). Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d). Color orthophotography of 2009 from Regional Ministry of the Environment.

@@

**Figure 2.** ***a)*** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error. See main text for details on EVI calculation. ***b)*** Percentage of pixels showing browning, greening or no changes during the 2005 and 2012 drought events according to EVI standardized anomalies. See main text for an explanation of greening and browning.

@@

**Figure 3.** Basal Area Increment (BAI) chronologies of *Q. pyrenaica* for northern population (SJ; *red*) and southern ones: low-elevation (CA-Low; *green*) and high-elevation (CA-High, *blue*) sites. Shading areas correspond to standard error of the mean. Number of series is displayed in the upper plot. Only years replicated with # series > 5 are shown. Linear trends since 1975 are indicated for all sites (numbers indicate $r^2$ values; asterisks indicate significant linear trend, *p < 0.001*).

@@

**Figure 4.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50% of GC (see Material and methods). Note that y-axes do not correspond in all of the three panels for the sake of clarity. Error bars indicate standard error.

@@

**Figure 5.** Resilience metrics of tree-growth for eight severe drought events since 1950 (see main text for details) as a function of drought severity. *Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*). Points indicate resilience metrics for oak populations: SJ (*red*), CA-High (*blue*) and CA-Low (*green*). Resilience metrics were computed for each population (sample depth > 10) and drought event. Gray lines represent overall relationships for each Resilience metrics.

@@

**Figure 6.** Comparison of the response of  *Q. pyrenaica* forests to drought in terms of resistance, recovery, and resilience of greenness (a) and tree growth (b). For EVI, northern populations (*red circle*) were compared with southern ones (*green circle*). For BAI, the more xeric northern population (San Juan, SJ; *red circle*) was compared with the two southern populations, Cáñar-High (CA-High; *blue circle*) and Cáñar-Low (CA-Low; *green circle*). Different letters indicate significant *post hoc* differences between groups (see Material and methods for details).

@@

**Figure 7.** Correlation coefficients found by relating tree-ring residual chronologies (RWI) of *Q. pyrenaica* and monthly climatic data: precipitation and 6-month SPEI (a), minimum (b) and maximum (c) temperatures. *green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation southern site (CA-High). Asterisks indicate significant ($P < 0.05$) correlation coefficients.

@@

**Figure S1.** ***a)*** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017. Points represent the mean, and error bars the standard error. The black line indicates mean for the entire period (585 mm). The red lines represent -1 and -2 standard deviation (dotted and dashed lines, respectively). The blue lines represent +1 and +2 standard deviation (dotted and dashed lines, respectively). Years with average values below -1SD are labeled. Data from 28 meteorological stations distributed around the Sierra Nevada area (from the National Spanish Meteorological Services, AEMET). ***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (blue line) and 2011-2012 (red line). The boxplot representing the average from 1950-2015 period. Data from meteorological station Granada, Base Aérea. ***b)*** Drought severity in Sierra Nevada for the 1901-2016 period based on the Standardized Precipitation-Evapotranspiration Index (SPEI). Data from Global SPEI database (http://spei.csic.es/database.html). We took the SPEI data for a 12-month scale and for all 0.5º grid cells covering Sierra Nevada. Horizontal gray bars indicate the years 2005 and 2012.

@@

**Figure S2.** ***a)*** Residual tree-ring chronologies determined for the *Q. pyrenaica* sites. Dashed red lines indicate the start of the reliable period (EPS > 0.85). Dotted black lines show the severe drought years identified in our climatic data (Table S3 and Figure S1). ***b)*** Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (right-axis). Data for number of trees > 2 is shown.

@@

**Figure S3.** ***a)*** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time domains after pre-filtering the time series with increasing size of the moving-average window (1 to 40 years). Each site chronology was smoothed using centered moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair of chronologies was calculated. Significance was tested using 1000 bootstrap replicates and with 95% confidence intervals built using the R package `boot`. ***b)*** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).

@@

##### 
**Figure S4.** EVI annual profile (average of the period 2000-2016) for *Q. pyrenaica* forests in Sierra Nevada and drought events. Horizontal bars correspond to the most severe droughts for Sierra Nevada since 1900 (computed as in Table S3). Their position indicates the start and end months of each drought event. Bars lengths show the duration of the drought event (number of consecutive months with SPEI lower than -1.28, see [@Pascoa2017].

@@

**Figure S5.** Resilience metrics of the tree growth for severe drought events since 1950 (excluding 1995 drought event). *Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*). Points indicate resilience metrics for oak populations: SJ (*blue*), CA-High (*red*) and CA-Low (*green*). Resilience metrics were computed for each population (sample depth > 10) and drought event. The gray line represents overall relationship for each Resilience metrics.

### 2019-12-10 man/tables/t_dendro_summary.Rmd
update ms

**Table 2.** Characteristics of the mean tree ring chronologies. Values of length between parentheses indicate number of years replicated with more than five series. RW = mean annual ring width (standard deviation in parenthesis). MS = mean sensitivity. AR(1) = mean autocorrelation of raw series. Rbt = mean correlation between series. EPS = mean expressed population signal. EPS and Rbt were calculated for the mean residual chronologies of growth indices.

>>>>

**Table 2.** Characteristics of the mean tree-ring chronologies. Length values in parentheses indicate the number of years replicated with more than five series. RW = mean annual ring width (standard deviation in parenthesis). MS = mean sensitivity. AR(1) = mean autocorrelation of raw series. Rbt = mean correlation between series. EPS = mean expressed population signal. EPS and Rbt were calculated for the mean residual chronologies of growth indices.

### 2019-12-10 man/tables/t_droughts.Rmd
update review GGI
update ms

##### 
**Table S3.** Drought events for the 1901-2016 period for Sierra Nevada based on SPEI index (12 months scale). See methods for details.

>>>>

##### 
**Table S3.** Drought events for the 1901-2016 period for Sierra Nevada ranked according to drought severity calculated from the SPEI index (12 months scale). See Material and methods for details.

### 2019-12-10 man/tables/t_forest_history.Rmd
update ms

**Table S4.** Review of the forest and management history of the samplin site. An exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest. Several documentary sources were reviewed: historical documents and maps [*e.g.* @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2016]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@JimenezOlivencia2015; @MorenoLlorca2016].

>>>>

**Table S4.** Review of the forest and management history of the sampling sites. Historical documents were exhaustively reviewed to compile information on socio-economical activities affecting forests: historical documents and maps [*e.g.* @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information on recent wildfires events and forest-management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2016]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@JimenezOlivencia2015; @MorenoLlorca2016].

### 2019-12-10 man/tables/t_huber_res.Rmd
update ms

**Table S2.** Robust measures of central tendency of resilience indices for greenness (EVI) and tree-growth (BAI), grouped by drought events, site and interaction. Measures of central tendency are M-estimators based on Huber's Psi (see material and methods). 95 % confidence intervals using 3000 bootstrap are included between parentheses. *Total* corresponds to the average of 2005 and 2012.

>>>>

**Table S2.** Robust measures of central tendency of resilience indices for greenness (EVI) and tree growth (BAI), grouped by drought events, site, and interaction. Measures of central tendency are M-estimators based on Huber's Psi (see Material and methods).In parentheses are the 95% confidence intervals using 3000 bootstraps. *Total* corresponds to the average of 2005 and 2012.

### 2019-12-10 man/tables/t_robust_anovas.Rmd
update ms

**Table 3.** Robust two-way ANOVAs of the resilience metrics of greenness (EVI) and tree-growth (BAI) for the two drought events (in 2005 and 2012) and site.

>>>>

**Table S1.** Robust two-way ANOVAs of the resilience metrics of greenness (EVI) and tree growth (BAI) for the two drought events (in 2005 and 2012) and site.

### 2019-12-10 man/tables/t_sampled_plots.Rmd
update ms

**Table 1.** Characteristics of sampled plot. Lat = latitude; Long = longitude. Dbh and height of all trees, Basal Area (BA), Density and SRD (Size ratio proportional to distance) are computed for all trees within a 10-m radius of focal trees (see methods). Temp.: annual average of mean monthly minimun and maximum temperatures. Values shown here correspond to site averages. Standard deviations are shown in parentheses. Different letters indicate statistically significant differences between sites (Kruskal-Wallis test followed by Dunn’s test, p<0.05). Stands were monospecific, hence all results correspond to oak data.

>>>>

**Table 1.** Characteristics of sampled plot. Lat = latitude; Long = longitude. Dbh and height of all trees, Basal Area (BA), Density and SRD (size ratio proportional to distance) are computed for all trees within a 10-m radius of focal trees (see Material and methods). Temp.: annual average of mean monthly minimun and maximum temperatures. Values shown here correspond to site averages. Standard deviations are shown in parentheses. Different letters indicate statistically significant differences between sites (Kruskal-Wallis test followed by Dunn’s test, p<0.05). Stands were monospecific, hence all results correspond to oak data.
